Star-galaxy Classification Using Convolutional Neural Networks

### 2016-05-17 paper/main.tex
update paper

The goal of learning is to find a set of weights and biases such that
the output from the network $\mathbf{y}=\left\{ y_1, y_2, \dots, y_n \right\}$
approximates the desired output
$\hat{\mathbf{y}}=\left\{\hat{y}_1, \hat{y}_2, \dots, \hat{y}_n \right\}$
as closely as possible for all input $\mathbf{x}$.
We can formulate this as the minimization of a loss function $L(\mathbf{x},\hat{\mathbf{y}})$
over the training data.
In this work, we use \textit{cross-entropy},
\begin{equation}
  L(y, \hat{y}) = -\hat{y} \log_2 y - (1 - \hat{y}) \log_2(1 - y),
\end{equation}
where $\hat{y}$ is the actual truth value (\eg 0 or 1)
and $y$ is the probability prediction made by the model.
We compute the loss function by taking the average of all cross-entropies in the sample.
Given $N$ samples, the loss function becomes
\begin{equation}
  L(y, \hat{y}) = - \frac{1}{N} \sum_{i=1}^{N} \hat{y}_i  \log_2 y_i
    + (1 - \hat{y}_i)  \log_2 (1 - \hat y_i).
\end{equation}
where $\hat{y}_i$ is the truth value and probability estimate of the $i$-th data,
respectively.

To find the weights $\mathbf{w}$ and biases $\mathbf{b}$ which minimize the loss,
we use a technique called \textit{gradient descent},
where we the following rules to update the parameters in each layer $j$:
\begin{align}
  \mathbf{w}_j &\rightarrow
  \mathbf{w}^{\prime}
  = \mathbf{w}_j - \eta \frac{\partial L}{\partial \mathbf{w}_j} \\
  \mathbf{b}_j &\rightarrow
  \mathbf{b}^{\prime}
  = \mathbf{b}_j - \eta \frac{\partial L}{\partial \mathbf{b}_j},
  \label{eq:gradeitn_descent}
\end{align}
where $\eta$ is a small, positive number known as the \textit{learning rate}.
A common approach to speed up training is to split the training data
into mini-batches and compute the gradients in
Equation~\ref{eq:gradeitn_descent} for each mini-batch separately instead of
the entire training data.
This results in much faster convergence, as training examples are usually correlated
and the gradient computed from a randomly chosen mini-batch is a good approximation
of the overall gradient.
We define an \textit{epoch} as a single, complete pass through the training data,
and full training usually requires many epochs.
At the end of each epoch, we evaluate the loss function on the validation set,
and the model that minimizes the loss is chosen as the best model.

@@

One of the advantages of a probabilistic classification is
that the threshold can be adjusted to produce
a more complete but less pure sample,
or a less complete but more pure one.
We can compare the performance of different classification techniques
by assuming an arbitrary operating condition.
For example, weak lensing science measurements
of the DES require $c_g > 0.960$ and $p_g > 0.778$
to control both the statistical and systematic errors
on the cosmological parameters,
and $c_s > 0.250$ and $p_s > 0.970$
for stellar Point Spread Function (PSF) calibration
\citep{soumagnac2013star}.
Although these values will likely be different
for the science cases of the CFHTLenS data,
we adopt these values to compare the classification performance
at a reasonable operating condition.
Thus, we compute $p_{g}$ at $c_g=0.960$
and $c_{s}$ at $p_s=0.250$.

@@

A fully probabilistic classifier predicts not only the class label,
but also its confidence level on the prediction.
In a well-calibrated classifier, the posterior class probability estimates
should coincide with the proportion of objects that truly belong to a certain class.
Probability \textit{calibration curves}~\citep[or reliability curves;][]{degroot1983comparison}
are often used to display this relationship,
where we bin the probability estimates and plot the
fraction of positive examples versus the predicted probability in each bin
(see Figures~\ref{fig:clens_calibration_conv} and fig:\textcolor{red}{xxx}).

@@

Ideally, the probabilistic output of a classifier would be used in subsequent
scientific analyses.
For example, one can weight each object by the probability that it is a galaxy
when measuring auto-correlation functions of luminous
galaxies~\citep{ross2011ameliorating}.
In other words, given a well-calibrated classifier,
instead of counting each galaxy equally, a galaxy could be
counted as $P_{\rmn{class}}$, the probability estimate.
This should, in principle, remove the contamination effect of stars.
For a perfect classifier, we can count the number of galaxies by summing
the values of $P_{\rmn{class}}$.
Thus, we measure the reliability of the classifier output by the absolute error
in the estimation of number of galaxies,
\begin{equation}
  \frac{\mid \Delta N_g \mid}{N_g} =
  \frac{\big| N_g - \sum_{i=1}^{N} P_{\rmn{class}} \big|}{N_g}
\end{equation}

@@

\begin{figure*}
  \centering
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/clens_calibration_conv.pdf}
    \caption{
      Calibration curve for ConvNet as applied to the CFHTLenS data set.
      $p_{\rmn{gal}}$ is the fraction of objects that are galaxies.
      $p_{\rmn{conv}}$ is the probabilistic output generated by ConvNet.
      The dashed line displays the relationship
      $p_{\rmn{gal}} = p_{\rmn{conv}}$.
      The $1 \sigma$ error bars are computed following the method
      of \citet{paterno2004calculating}.
      }
    \label{fig:clens_calibration_conv}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figures/clens_calibration_tpc.pdf}
    \caption{Calibration curve for TPC as applied to the CFHTLenS data set.}
    \label{fig:clens_calibration_tpc}
  \end{minipage}
\end{figure*}

@@

In Figure~\ref{fig:clens_mag}, we compare the galaxy completeness
and star purity values for ConvNet and TPC as functions of $i$-band magnitude
for the differential counts.
We use the kernel density estimation~\citep[KDE][]{silverman1986density}
with the Gaussian kernel.
As the first panel shows, KDE is able to smooth the fluctutions in the distribution
without binning.
While ConvNet shows a slightly better performance than $\rmn{TPC}_{\rmn{morph}}$
in galaxy purity, the star completeness plots for ConvNet and TPC
are nearly identical.
Again, $\rmn{TPC}_{\rmn{phot}}$ performs significantly worse than both
ConvNet and TPC, and this suggests that ConvNets are able to learn
the shape information automatically from the images.
We note that, at these operating conditions ($c_g=0.96$, $p_s=0.97$),
both ConvNet and $\rmn{TPC}_{\rmn{morph}}$ outperform the
star-galaxy classification provided by
the CFHTLenS pipeline~\cite{hildebrandt2012cfhtlens} across all magnitudes.
We do not show the CFHTLenS classifications to avoid cluttering the plots.

@@

We also show the galaxy purity and star completeness values as functions of
$g-r$ color in Figure~\ref{fig:clens_g_r}.
$\rmn{TPC}_{\rmn{morph}}$ provides slightly better
completeness and purity than ConvNet between $0.8 \la g-r \la 1.6$ while
ConvNet outperforms $\rmn{TPC}_{\rmn{morph}}$ or is similar to
$\rmn{TPC}_{\rmn{morph}}$ in the remaining regions.

Figure~\ref{fig:clens_calibration_conv} shows the calibration curve
that compares $P_{\rmn{gal}}$,
the fraction of objects that are galaxies (as determined from their spectra)
to $P_{\rmn{conv}}$ the probabilistic output that our ConvNet model produces.
The calibration curve is nearly diagonal, which implies that ConvNet
is well-calibrated and we can treat $P_{\rmn{conv}}$ as the probability
that an object is a galaxy.
Figure~\ref{fig:clens_calibration_tpc} shows the calibration curve
for the probabilistic output of TPC,
which is apparently not as well-calibrated as ConvNet.
These calibration curves visually confirm the results in
Table~\ref{table:clens_metrics} that the calibration error
of ConvNet is about 20\% lower than that of TPC.
While probabilistic predictions can be further calibrated by using,
\eg isotonic calibration~\citep{zadrozny2001obtaining},
we do not consider additional probability calibration in this work.

@@

When highly optimized features (\eg the $\rmn{concentration}$ of SDSS)
are not readily available from the catalog, a ConvNet model becomes more
attractive.

@@

\begin{figure*}
  \centering
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/sdss_calibration_conv.pdf}
    \caption{Calibration curve for ConvNet as applied to the SDSS data set.}
    \label{fig:sdss_calibration_conv}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/sdss_calibration_tpc.pdf}
    \caption{Calibration curve for TPC as applied to the CFHTLenS data set.}
    \label{fig:sdss_calibration_tpc}
  \end{minipage}
\end{figure*}

@@

In Figure~\ref{fig:sdss_mag}, we compare the galaxy completeness
and star purity values for ConvNet and TPC as functions of $i$-band magnitude
for the differential counts.
We note that $\rmn{TPC}_{\rmn{morph}}$ outperforms the star-galaxy
classifier used by the SDSS pipeline (\ie an object is classifed as a galaxy
if the concentration parameter is greater than 0.145) across all magnitudes.
We do not show the SDSS classifications to avoid cluttering the plots.
While ConvNet shows a similar but slighly worse performance than
$\rmn{TPC}_{\rmn{morph}}$,
the galaxy purity and star completeness values of ConvNet begin to
drop at faint magnitudes $i \la 21$.
Again, $\rmn{TPC}_{\rmn{phot}}$ performs significantly worse than both
ConvNet and TPC at bright magnitudes.
One reason that ConvNet fails to outperform $\rmn{TPC}_{\rmn{phot}}$,
especially at faint magnitudes,
might be its over-reliance on morphological features.
Near a survey's limit, the measurement uncertainties generally increase,
and morphology is not a reliable metric for star-galaxy classification.

@@

We also show the galaxy purity and star completeness values as functions of
$g-r$ color in Figure~\ref{fig:sdss_g_r}.
$\rmn{TPC}_{\rmn{morph}}$ provides slightly better
completeness and purity than ConvNet between $0.8 \la g-r \la 1.6$ while
ConvNet outperforms $\rmn{TPC}_{\rmn{morph}}$ or is similar to
$\rmn{TPC}_{\rmn{morph}}$ in the remaining regions.

Figure~\ref{fig:sdss_calibration_conv} shows the calibration curve
that compares $P_{\rmn{gal}}$,
the fraction of objects that are galaxies (as determined from their spectra)
to $P_{\rmn{conv}}$ the probabilistic output that our ConvNet model produces.
The calibration curve is nearly diagonal, which implies that ConvNet
is well-calibrated and we can treat $P_{\rmn{conv}}$ as the probability
that an object is a galaxy.
Figure~\ref{fig:sdss_calibration_tpc} shows the calibration curve
for the probabilistic output of TPC,
which is apparently not as well-calibrated as ConvNet.
These calibration curves visually confirm the results in
Table~\ref{table:sdss_metrics} that the calibration error
of ConvNet is about 20\% lower than that of TPC.

>>>>

The goal of learning is to find a set of weights and biases such that
the output from the network $\mathbf{y}=\left\{ y_1, y_2, \dots, y_n \right\}$
approximates the desired output
$\hat{\mathbf{y}}=\left\{\hat{y}_1, \hat{y}_2, \dots, \hat{y}_n \right\}$
as closely as possible for all input $\mathbf{x}$.
We can formulate this as the minimization of a loss function $L(\mathbf{x},\hat{\mathbf{y}})$
over the training data.
In this work, we use \textit{cross-entropy},
\begin{equation}
  L(y, \hat{y}) = -\hat{y} \log_2 y - (1 - \hat{y}) \log_2(1 - y),
\end{equation}
where $\hat{y}$ is the actual truth value (\eg 0 or 1)
and $y$ is the probability prediction made by the model.
We compute the loss function by taking the average of all cross-entropies in the sample.
Given $N$ samples, the loss function becomes
\begin{equation}
  L(y, \hat{y}) = - \frac{1}{N} \sum_{i=1}^{N} \hat{y}_i  \log_2 y_i
    + (1 - \hat{y}_i)  \log_2 (1 - \hat y_i).
\end{equation}
where $\hat{y}_i$ and $y_i$ are the truth value and probability estimate of
the $i$-th data, respectively.

To find the weights $\mathbf{w}$ and biases $\mathbf{b}$ which minimize the loss,
we use a technique called \textit{gradient descent},
where we use the following rules to update the parameters in each layer $j$:
\begin{align}
  \mathbf{w}_j &\rightarrow
  \mathbf{w}^{\prime}
  = \mathbf{w}_j - \eta \frac{\partial L}{\partial \mathbf{w}_j} \\
  \mathbf{b}_j &\rightarrow
  \mathbf{b}^{\prime}
  = \mathbf{b}_j - \eta \frac{\partial L}{\partial \mathbf{b}_j},
  \label{eq:gradeitn_descent}
\end{align}
where $\eta$ is a small, positive number known as the \textit{learning rate}.
A common approach to speed up training is to split the training data
into mini-batches and compute the gradients in
Equation~\ref{eq:gradeitn_descent} for each mini-batch separately instead of
the entire training data.
This results in much faster convergence, as training examples are usually correlated
and the gradient computed from a randomly chosen mini-batch is a good approximation
of the overall gradient.
We define an \textit{epoch} as a single, complete pass through the training data,
and full training usually requires many epochs.
At the end of each epoch, we evaluate the loss function on the validation set,
and the model that minimizes the validation loss is chosen as the best model.

@@

One of the advantages of a probabilistic classification is
that the threshold can be adjusted to produce
a more complete but less pure sample,
or a less complete but more pure one.
We can compare the performance of different classification techniques
by assuming an arbitrary operating condition.
For example, weak lensing science measurements
of the DES require $c_g > 0.960$ and $p_g > 0.778$
to control both the statistical and systematic errors
on the cosmological parameters,
and $c_s > 0.250$ and $p_s > 0.970$
for stellar Point Spread Function (PSF) calibration
\citep{soumagnac2013star}.
Although these values will likely be different
for the science cases of the CFHTLenS data,
we adopt these values to compare the classification performance
at a reasonable operating condition.
Thus, we compute $p_{g}$ at $c_g=0.960$
and $c_{s}$ at $p_s=0.970$.

@@

A fully probabilistic classifier predicts not only the class label,
but also its confidence level on the prediction.
In a well-calibrated classifier, the posterior class probability estimates
should coincide with the proportion of objects that truly belong to a certain class.
Probability \textit{calibration curves}~\citep[or reliability curves;][]{degroot1983comparison}
are often used to display this relationship,
where we bin the probability estimates and plot the
fraction of positive examples versus the predicted probability in each bin
(see Figures~\ref{fig:clens_calibration_conv} and \ref{fig:clens_calibration_tpc}).

@@

Ideally, the probabilistic output of a classifier would be used in subsequent
scientific analyses.
For example, one can weight each object by the probability that it is a galaxy
when measuring auto-correlation functions of luminous
galaxies~\citep{ross2011ameliorating}.
In other words, given a well-calibrated classifier,
instead of counting each galaxy equally, a galaxy could be
counted as $P_{\rmn{class}}$, the probability estimate.
This should, in principle, remove the contamination effect of stars.
For a perfect classifier, we can count the number of galaxies by summing
the values of $P_{\rmn{class}}$.
Thus, we measure the reliability of the classifier output by the absolute error
in the estimation of number of galaxies,
\begin{equation}
  \frac{\mid \Delta N_g \mid}{N_g} =
  \frac{\big| N_g - \sum_{i=1}^{N} P_{\rmn{class}} \big|}{N_g}.
\end{equation}

@@

\begin{figure*}
  \centering
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/clens_calibration_conv.pdf}
    \caption{
      Calibration curve for ConvNet as applied to the CFHTLenS data set.
      $p_{\rmn{gal}}$ is the fraction of objects that are galaxies.
      $p_{\rmn{conv}}$ is the probabilistic output generated by ConvNet.
      The dashed line displays the relationship
      $p_{\rmn{gal}} = p_{\rmn{conv}}$.
      The $1 \sigma$ error bars are computed following the method
      of \citet{paterno2004calculating}.
      }
    \label{fig:clens_calibration_conv}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figures/clens_calibration_tpc.pdf}
    \caption{Calibration curve for $\rmn{TPC}_{\rmn{morph}}$
      as applied to the CFHTLenS data set.}
    \label{fig:clens_calibration_tpc}
  \end{minipage}
\end{figure*}

@@

In Figure~\ref{fig:clens_mag}, we compare the galaxy completeness
and star purity values for ConvNet and TPC as functions of $i$-band magnitude
for the differential counts.
We use the kernel density estimation~\citep[KDE][]{silverman1986density}
with the Gaussian kernel.
As the first panel shows, KDE is able to smooth the fluctutions in the distribution
without binning.
While ConvNet shows a slightly better performance than $\rmn{TPC}_{\rmn{morph}}$
in galaxy purity, the star completeness plots for ConvNet and TPC
are nearly identical.
Again, $\rmn{TPC}_{\rmn{phot}}$ performs significantly worse than both
ConvNet and TPC, and this suggests that ConvNets are able to learn
the shape information automatically from the images.
We note that, at these operating conditions ($c_g=0.96$, $p_s=0.97$),
both ConvNet and $\rmn{TPC}_{\rmn{morph}}$ outperform the
star-galaxy classification provided by
the CFHTLenS pipeline~\citep{hildebrandt2012cfhtlens} over all magnitudes.

@@

We also show the galaxy purity and star completeness values as functions of
$g-r$ color in Figure~\ref{fig:clens_g_r}.
$\rmn{TPC}_{\rmn{morph}}$ provides slightly better
completeness and purity than ConvNet between $0.8 \la g-r \la 1.6$ while
ConvNet outperforms $\rmn{TPC}_{\rmn{morph}}$ in the remaining regions.

Figure~\ref{fig:clens_calibration_conv} shows the calibration curve
that compares $P_{\rmn{gal}}$,
the fraction of objects that are galaxies (as determined from their spectra),
to $P_{\rmn{conv}}$, the probabilistic output that our ConvNet model produces.
The calibration curve is nearly diagonal, which implies that ConvNet
is well-calibrated and we can treat $P_{\rmn{conv}}$ as the probability
that an object is a galaxy.
Figure~\ref{fig:clens_calibration_tpc} shows the calibration curve
for the probabilistic output of $\rmn{TPC}_{\rmn{morph}}$,
which is apparently not as well-calibrated as ConvNet.
These calibration curves visually confirm the results in
Table~\ref{table:clens_metrics} that the calibration error
of ConvNet is about 20\% lower than that of $\rmn{TPC}_{\rmn{morph}}$.
While probabilistic predictions can be further calibrated by using,
\eg isotonic calibration~\citep{zadrozny2001obtaining},
we do not consider additional probability calibration in this work.

@@

@@

\begin{figure*}
  \centering
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/sdss_calibration_conv.pdf}
    \caption{Calibration curve for ConvNet as applied to the SDSS data set.}
    \label{fig:sdss_calibration_conv}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/sdss_calibration_tpc.pdf}
    \caption{Calibration curve for $\rmn{TPC}_{\rmn{morph}}$
      as applied to the SDSS data set.}
    \label{fig:sdss_calibration_tpc}
  \end{minipage}
\end{figure*}

@@

In Figure~\ref{fig:sdss_mag}, we compare the galaxy completeness
and star purity values for ConvNet and TPC as functions of $i$-band magnitude
for the differential counts.
We note that $\rmn{TPC}_{\rmn{morph}}$ outperforms the star-galaxy
classifier used by the SDSS pipeline (\ie an object is classifed as a galaxy
if the concentration parameter is greater than 0.145) over all magnitudes.
We do not show the SDSS classifications to avoid cluttering the plots.
While ConvNet shows a similar but slighly worse performance than
$\rmn{TPC}_{\rmn{morph}}$,
the galaxy purity and star completeness values of ConvNet begin to
drop at faint magnitudes $i \la 21$.
Again, $\rmn{TPC}_{\rmn{phot}}$ performs significantly worse than both
ConvNet and TPC at bright magnitudes.
One reason that ConvNet fails to outperform $\rmn{TPC}_{\rmn{phot}}$,
especially at faint magnitudes,
might be its over-reliance on morphological features.
Near a survey's limit, the measurement uncertainties generally increase,
and morphology is not a reliable metric for star-galaxy classification.

@@

We also show the galaxy purity and star completeness values as functions of
$g-r$ color in Figure~\ref{fig:sdss_g_r}.
ConvNet performs slightly better than $\rmn{TPC}_{\rmn{morph}}$ in both
galaxy completeness and star purity between $0.7 \la g-r \la 2.0$
where the stellar fraction is relatively low.
Both $\rmn{TPC}_{\rmn{morph}}$ and $\rmn{TPC}_{\rmn{phot}}$ outperform
ConvNet in the region $g-r \la 0.8$ where the stellar fraction is higher.

Figure~\ref{fig:sdss_calibration_conv} and \ref{fig:sdss_calibratoin_tpc}
show the calibration curves of ConvNet and $\rmn{TPC}_{\rmn{morph}}$,
respectively.
Compared to Figure~\ref{fig:clens_calibration_conv}, our ConvNet as applied to
the SDSS data set is not as well-calibrated as when it is applied to the
CFHTLenS data set.
